Abstractions for Multi-Sorted Substitutions

### 2023-08-25 paper/paper.tex
Update.

@@

  We identify two drawbacks with this style of syntax:
  \begin{enumerate}
  \item the syntactic categories (\texttt{Kind}, \texttt{Type}, and \texttt{Expr})
    have different types, which makes it difficult to treat them uniformly; and
  \item the different sorts of variables are modeled separately, which requires to define not just
    type-in-type and expression-in-expression substitution, but also type-in-expression substitution.
    Consequently, interaction lemmas between those substitutions are required.
    In general, this leads to a combinatory explosion of
    substitutions, e.g.\ adding kind variables would lead to a total of 6
    different substitutions.
  \end{enumerate}

  \subsection{Multi-Sorted Syntax}
  To avoid these drawbacks, we define the syntax as a single
  sort-indexed type of terms.

@@

  As we intend to have exactly two kit instances, we choose names of
  the form \texttt{x/y}, where \texttt{x} is the name we choose for the
  variable instance, and \texttt{y} is the name we choose for the term
  instance.
  For example the parameter type \texttt{\_∋/⊢\_} will be instanciated
  to \texttt{\_∋\_} for the variable kit, and to \texttt{\_⊢\_} for the
  term kit.

  A kit provides the following operations:
  \begin{itemize}
  \item 
    \texttt{id/`} converts a variable \texttt{S ∋ s} into a
    \texttt{S ∋/⊢ s}. For the variable kit, \texttt{S ∋/⊢ s} is instanciated to \texttt{S ∋ s}, so this
    operation is the identity. For the term kit, \texttt{S ∋/⊢ s} is
    instanciated to \texttt{S ⊢ s}, so this operation is the
    variable constructor \texttt{`\_}.
  \item 
    \texttt{`/id} converts a \texttt{S ∋/⊢ s} into a term \texttt{S ⊢ s}
    and is dual to the \texttt{id/`} operation.
  \item 
    \texttt{wk} shifts the DeBruin-Indices in a \texttt{S ∋/⊢ s}.
    The new, unused variable \texttt{zero} can assume any sort \texttt{s'}.
    For variables \texttt{wk} is the successor \texttt{suc}.
    For terms \texttt{wk} means applying a shifting renaming to the term.
  \end{itemize}

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \texttt{`/`-is-`} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \texttt{`\_}.
  \item \texttt{`/id-injective} and \texttt{id/`-injective} state that
    \texttt{`/id} and \texttt{id/`} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \texttt{wk-id/`} characterizes the behaviour of the \texttt{wk} function.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \texttt{S₁ →ᵣ S₂} and
    substitutions \texttt{S₁ →ₛ S₂}, which differ only in whether they
    return variables or a terms:
    \LibCode\KMap
    We call a value of type \texttt{S₁ →ₖ S₂} a \emph{map}.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \texttt{wk} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \texttt{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \texttt{zero} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \texttt{wk}:
    \LibCode\KWeaken
  \end{itemize}

>>>>

\newcommand*\ACode[1]{\AgdaFontStyle{#1}}
\newcommand*\AField[1]{\AgdaField{#1}}
\newcommand*\ACon[1]{\AgdaInductiveConstructor{#1}}
\newcommand*\AKw[1]{\AgdaKeyword{#1}}
\newcommand*\ADef[1]{\AgdaFunction{#1}}

@@

  We identify two drawbacks with this style of syntax:
  \begin{enumerate}
  \item the syntactic categories (\texttt{Kind}, \texttt{Type}, and \texttt{Expr})
    have different types, which makes it difficult to treat them uniformly; and
  \item the different sorts of variables are modeled separately, which requires to define not just
    type-in-type and expression-in-expression substitution, but also type-in-expression substitution.
    Consequently, interaction lemmas between those substitutions are required.
    In general, this leads to a combinatory explosion of
    substitutions, e.g.\ adding kind variables would lead to a total of 6
    different substitutions.
  \end{enumerate}
  To avoid these drawbacks, we instead use a multi-sorted syntax.

  \subsection{Multi-Sorted Syntax}
  A multi-sorted syntax is defined by a single type of sort-indexed terms.

@@

  As we intend to have exactly two kit instances, we choose names of
  the form \texttt{x/y}, where \texttt{x} is the name we choose for the
  variable instance, and \texttt{y} is the name we choose for the term
  instance.
  For example the parameter type \texttt{\_∋/⊢\_} will be instanciated
  to \ACon{\_∋\_} for the variable kit, and to \AField{\_⊢\_} for the
  term kit.

  A kit provides the following operations:
  \begin{itemize}
  \item 
    \AField{id/`} converts a variable \ACode{S \ACon{∋} s} into a
    \ACode{S \AField{∋/⊢} s}. For the variable kit, \ACode{S ∋/⊢ s} is instanciated to \ACode{S \ACon{∋} s}, so this
    operation is the identity. For the term kit, \ACode{S ∋/⊢ s} is
    instanciated to \ACode{S \AField{⊢} s}, so this operation is the
    variable constructor \AField{`\_}.
  \item 
    \AField{`/id} converts a \ACode{S ∋/⊢ s} into a term \ACode{S \AField{⊢} s}
    and is dual to the \AField{id/`} operation.
  \item 
    \AField{wk} shifts the DeBruin-Indices in a \ACode{S ∋/⊢ s}.
    The new, unused variable \ACon{zero} can assume any sort \ACode{s'}.
    For variables \AField{wk} is the successor \ACon{suc}.
    For terms \AField{wk} means applying a shifting renaming to the term.
  \end{itemize}

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \AField{`/`-is-`} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \ACode{`\_}.
  \item \AField{`/id-injective} and \AField{id/`-injective} state that
    \AField{`/id} and \AField{id/`} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \AField{wk-id/`} characterizes the behaviour of the \AField{wk} function.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \texttt{S₁ →ᵣ S₂} and
    substitutions \texttt{S₁ →ₛ S₂}, which differ only in whether they
    return variables or a terms:
    \LibCode\KMap
    We call a value of type \texttt{S₁ →ₖ S₂} a \emph{map}.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \texttt{wk} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \texttt{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \texttt{zero} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \texttt{wk} on terms:
    \LibCode\KWeaken
  \end{itemize}
